---
title: "Accessing fishbase data"
output: 
  html_document:
    fig_height: 3
    fig_width: 4
    highlight: haddock
    theme: united
    toc: no
    toc_float: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


The **rfishbase**-package provides an interface to the [fishbase.org](http://fishbase.org) database. In this snipped we focus only on extracting the parameters of length-coefficient and von Bertalanffy growth curve.

```{r}
library(tidyverse)
library(rfishbase)
```

Lets just take one species as an example:
```{r}
latin <- c("Gadus morhua")
```

We can extract the length-weight parameters with the `length_weight`-function:
```{r}
lwcoeff <- length_weight(latin)
glimpse(lwcoeff)
```

We have 37 variables and 18 observation (the latter varies between species). The parameters we are interested in are **a** and **b** that describe the relationship between length and weight:

$L = aL^b$

```{r, echo = FALSE}
df <-
  lwcoeff %>% 
  select(sciname, a, b) %>% 
  drop_na() %>% 
  mutate(id = 1:n())

expand.grid(length = 0:110,
            id = df$id) %>% 
  as_tibble() %>% 
  left_join(df) %>% 
  mutate(weight = a * length^b / 1e3) %>% 
  ggplot(aes(length, weight, group = id)) +
  geom_line(alpha = 0.5)
```

We note that for a given length we observe quite different prediction in weight. Although the devil is in the detail, we proceed by just taking the median value:

```{r}
lwcoeff <-
  lwcoeff %>% 
  select(sciname, a, b) %>% 
  drop_na() %>% 
  group_by(sciname) %>% 
  summarise(n.lw = n(),
            a = median(a),
            b = median(b))
lwcoeff
```

We similarly can extract the von Bertalanffy parameters using the `popgrowth`-function:
```{r}
vbl <- popgrowth(latin)
glimpse(vbl)
```

Here we have 80 variables and 104 observations (again the latter varies between species). The parameters we are interested in are **Loo** ($L_{inf}$), **K** and **to** ($_{0}$) that describe the relationship between age and length:

$L = L_{inf} (1 - e^{-K (age - t_{0})})$

```{r, echo = FALSE}
df <-
  vbl %>% 
  select(sciname, Loo, K, to) %>% 
  drop_na() %>% 
  mutate(id = 1:n())

expand.grid(age = 0:20,
            id = df$id) %>% 
  as_tibble() %>% 
  left_join(df) %>% 
  mutate(length = Loo * (1 - exp(-(K * (age - to))))) %>% 
  ggplot(aes(age, length, group = id)) +
  geom_line(alpha = 0.2)
```

Here, there are substantial difference among the parameter sets, in this case we are even getting predictions that do not make any biological sense. Lets continue to be wreak-less and just take the median values:
```{r}
vbl <-
  vbl %>%
  select(sciname, Loo, K, to) %>% 
  drop_na() %>% 
  group_by(sciname) %>% 
  summarise(n.vbl = n(),
            Loo = median(Loo),
            K   = median(K),
            to  = median(to))
vbl
```

We can easily join the tables:
```{r}
lwcoeff %>% 
  left_join(vbl)
```


