---
title: "Programming in R"
output: 
  html_document:
    fig_height: 4
    fig_width: 8
    highlight: haddock
    theme: united
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
minke <- read_csv('data/minke.csv')
```


# Conditional statements

It is often so that we would like some parts of our code to run only if some condition is fulfilled
We use `if` sentences for that.

```{r,eval=FALSE}
if(condition){
  ## this runs if the condition(s) are fulfilled
}
```

As a dummy example 

```{r}
x <- 10
## check if x is greater than 0
if(x>0){
  print('x is positive')
}
```

Sometimes we would like to check for more than one condition. We can use a combination of `if` and `else` sentences for that

```{r,eval=FALSE}
if(condition_1){
  ## this runs if condition 1 is fulfilled,
  ## then we skip the else-sentences below
} else if(condition_2){
  ## this runs if condition 2 is fulfilled,
  ## then we skip the else-sentence below
} else{
  ## this runs of neither condition 1 or 2 are fulfilled
}

```

So to continue with the dummy example:

```{r}
x <- 10
## check if x > 10
if(x>10){
  print('x is greater than 10')
} else if(x>=0){
  print('x is less than 10 but greater than or equal to 0')
} else {
  print('x is negative')
}
```


It is often so that we only have two conditions and they need to be vectorised. We can use the `ifelse` function for these cases

```{r,eval=FALSE}
ifelse(condition,
       'this runs if the condition is fulfilled',
       'this runs if the condition is not fulfilled')
```


As an example we can replace NA’s from data using the `ifelse` function:

```{r}

x <- c(1,2,3,NA,4,5)
## find all missing entries and replace them
x <- ifelse(is.na(x),0,x)
x
```

# Loops

### `for`-loop

We use a `for`-loop if we need some part of our code to run more than one time and we know how often it should run. We use an index/counter that counts how often we enter the loop:

```{r,eval=FALSE}
for(index in vector){
  ## This code runs as often as the length of the vector
}

```

Most of the time we write for loops like this:
```{r,eval=FALSE}
for(i in 1:10){
  ## do some stuff
  print(i)
}
```


### `while` loops

While-loops are similar to for-loops, but they are used when we need some part of our code to run more
than one time but we dont know how often it sould run. 

```{r, eval=FALSE}

while(condition){
  ## This code runs while the condition(s) are fulfilled
}

```

As a dummy example we can calculate the mean length of minke whales using a `for`-loop

```{r}
## find the number of entries in the data
n <- nrow(minke)
## calculate the mean length old fashion way
ML <- 0
for(i in 1:n){
  ML <- ML + minke$length[i]
}
ML <- ML/n
ML
```



