---
title: "Points in polygons"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

A question was asked how to determine which lon and lat points of a dataframe fall within a spacial polygon. Here is a simple demonstration (using shapes-objects as defined for the sp-packages [may later try to do this for the shape-objects from the sf-package]:

1. Needed libraries (install if not already on your computer):

```{r}
library(sp)
library(rgdal)
library(tidyverse)
```

2. Download an example shapefile (here roundfish area of the NS-IBTS):
```{r}
download.file(url = "http://gis.ices.dk/Shapefiles/6951716_output.zip",
              destfile = "data/6951716_output.zip", mode = "wb")
unzip("data/6951716_output.zip", exdir = "data")
ns <- readOGR(dsn = "/net/hafkaldi/export/home/haf/einarhj/edu/tcrenv2017/data",
              layer = "Shapefile")
```

3. The dataframe that contains some lon's and lat's (here we just generate some random data):
```{r}
d <- data_frame(lon =    rnorm(n = 100, mean =    5, sd =   3),
                lat =    rnorm(n = 100, mean =   55, sd =   2),
                effort = rnorm(n = 100, mean = 1000, sd = 200))  # unit in minutes
```

4. We need a little helper function (actually taken from the gisland-packages, see: https://github.com/einarhjorleifsson/gisland/blob/master/R/geo_inside.R):
```{r}
#' @title Spatial location of datapoints
#'
#' @description Returns spatial attributes of coordinates (longitude and latitude)
#' given spatial polygons.
#'
#' @param lon A numerical vector
#' @param lat A numerical vector
#' @param map Normally a spatialPolygonDataFrame
#' @param variable The variable name, stored in the map attribute table to be
#' returned. If missing (default) only boolean vector is returned indicating
#' if coordinates are inside or outside any region.
#'
#' @export
#'
geo_inside <- function(lon, lat, map, variable) {

  # Here we could try to pass the stuff to geo_inside2
  # The problem is that it does not work if we have holes
  if(class(map) == "data.frame") {

    message("The map is a data.frame, tryp geo_inside2")
    return(NULL)

  }

  # deal with missing lons and lats
  df <- data.frame(long = lon, lat = lat)
  df$id <- 1:nrow(df)
  df.redux <- df[!is.na(df$lon),]

  x <- sp::SpatialPoints(data.frame(long = df.redux$long, lat = df.redux$lat))
  sp::proj4string(x) <- sp::proj4string(map)

  x <- sp::over(x, map)

  x$id <- df.redux$id

  x <-
    data.frame(id = df$id) %>%
    dplyr::left_join(x)

  if(!missing(variable)) {
    if(is.factor(x[, variable])) {
      return(as.character(x[,variable]))
    } else {
      return(x[, variable])
    }
  } else {

    # cumbersome stuff - clarify
    if(is.null(dim(x))) {
      x <- as.logical(x)
      x <- ifelse(is.na(x), FALSE, x)
      return(x)
    } else {
      return(!is.na(x[,1]))
    }
  }
}
```

5. And now we can run the following to obtain information on inside which roundfish region each point belongs to (returns NA if it does not belong to any):
```{r}
d <-
  d %>% 
  mutate(strata = geo_inside(lon, lat, ns, "AreaName"))
```

6. A little plot:

```{r}
d %>% 
  ggplot() +
  theme_bw() +
  geom_path(data = ns, aes(long, lat, group = group)) +
  geom_point(aes(lon, lat, colour = strata)) +
  coord_quickmap()
```



#### Preliminary experimentation via the sf-package:
```{r}
library(sf)
ns <- sf::st_read("data/Shapefile.dbf")
# turn the dataframe into a geometry
d <- 
  data_frame(lon =    rnorm(n = 100, mean =    5, sd =   3),
                lat =    rnorm(n = 100, mean =   55, sd =   2),
                effort = rnorm(n = 100, mean = 1000, sd = 200)) %>% 
  st_as_sf(coords = c("lon", "lat"),
           crs = 4326,
           remove = FALSE)
d <-
  d %>% 
  st_join(ns)
glimpse(d)
ggplot() +
  geom_sf(data = ns) +
  geom_point(data = d, aes(lon, lat, colour = AreaName), size = 1) +
  labs(x = NULL, y = NULL)
```

