---
title: "Importing and tidying"
output: 
  html_document:
    fig_height: 4
    fig_width: 8
    highlight: haddock
    theme: united
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```


```{r}
library(tidyverse)
```

# Preamble
___

... say something on the DATRAS data (use links) and something about the objectives of this exercise (what will you learn).

... make sure to emphasise that this is a case example and the intent is not dig deeply into caluclation of survey indices.


# Importing
___

One can obtain DATRAS data using the `getDATRAS` function from the [icesDatras](https://github.com/ices-tools-prod/icesDatras)-package. Here will limit what will be retrieved to the North Sea International Bottom Trawl Survev (NS-IBTS), for years 1975 to 2017 and quarter 1:

```{r, eval = FALSE}
# not run
raw_st <- icesDatras::getDATRAS(record = "HH",
                                survey = "NS-IBTS",
                                years = 1975:2017,
                                quarters = 1)
raw_le <- icesDatras::getDATRAS(record = "HL",
                                survey = "NS-IBTS",
                                years = 1975:2017,
                                quarters = 1)
raw_ag <- icesDatras::getDATRAS(record = "CA",
                                survey = "NS-IBTS",
                                years = 1975:2017,
                                quarters = 1)
```

Because this takes some ten's of minutes to retrieve one would not do this everytime one is starting an R-session. Hence, normally one would do retrival steps ("downloading") **once** and save the data for more quicker retrieval "next time around".

In addition to the above we need some additional information.

Firstly we need to obtain the species:
```{r, eval = FALSE}
species <- 
  raw_le %>% 
  select(speccodetype = SpecCodeType, speccode = SpecCode) %>% 
  distinct() %>% 
  mutate(speccode = as.integer(speccode)) %>% 
  filter(!is.na(speccode))
tsn <- 
  species %>% 
  filter(speccodetype == "T")
out.tsn <- list()
for(i in 1:nrow(tsn)) {
  out.tsn[[i]] <- paste0("https://datras.ices.dk/WebServices/DATRASWebService.asmx/getSpecies?codename=tsn&code=",
                tsn$speccode[i]) %>% 
    icesDatras:::readDatras() %>% 
    icesDatras:::parseDatras()
}

aphia <- 
  species %>% 
  filter(speccodetype == "W")

out.aphia <- list()
for(i in 1:nrow(aphia)) {
  out.aphia[[i]] <- 
    paste0("https://datras.ices.dk/WebServices/DATRASWebService.asmx/getSpecies?codename=aphia&code=",
                aphia$speccode[i]) %>% 
    icesDatras:::readDatras() %>% 
    icesDatras:::parseDatras()
}

species2 <-
  bind_rows(out.tsn) %>% 
  bind_rows(bind_rows(out.aphia)) %>% 
  as_tibble() %>% 
  select(aphia, tsn, latin = latinname) %>% 
  distinct() %>% 
  gather(speccodetype, speccode, aphia:tsn) %>% 
  mutate(speccodetype = ifelse(speccodetype == "aphia", "W", "T")) %>% 
  filter(speccode > 0,
         !is.na(speccode)) %>% 
  mutate(latin = ifelse(latin == "Torpedo (Torpedo) marmorata", "Torpedo marmorata", latin)) %>% 
  distinct()

raw_le.tmp <- raw_le
colnames(raw_le.tmp) <- tolower(colnames(raw_le.tmp))
raw_le.tmp <-
  raw_le.tmp %>% 
  left_join(species2)
nrow(raw_le) - nrow(raw_le.tmp)


valid_aphia <- 
  raw_le.tmp %>% 
  filter(is.na(latin)) %>% 
  select(valid_aphia) %>% 
  mutate(valid_aphia = as.integer(valid_aphia),
         speccodetype = "W") %>% 
  distinct() %>% 
  drop_na()

out.aphia2 <- list()
for(i in 1:nrow(valid_aphia)) {
  out.aphia2[[i]] <- 
    paste0("https://datras.ices.dk/WebServices/DATRASWebService.asmx/getSpecies?codename=aphia&code=",
                valid_aphia$valid_aphia[i]) %>% 
    icesDatras:::readDatras() %>% 
    icesDatras:::parseDatras()
}

species3 <-
  out.aphia2 %>% 
  bind_rows() %>% 
  as_tibble() %>% 
  select(valid_aphia = aphia, latin = latinname) %>%
  distinct()

x.defined   <- raw_le.tmp %>% filter(!is.na(latin))
x.undefined <- raw_le.tmp %>% filter( is.na(latin)) %>% 
  select(-latin) %>% 
  left_join(species3)
x <- 
  bind_rows(x.defined, x.undefined)
x  %>% filter(is.na(latin)) %>% 
  group_by(year) %>% 
  count() %>% as.data.frame()
x %>% filter(is.na(latin)) %>% select(speccodetype, speccode, valid_aphia) %>% as_tibble() %>% drop_na() %>% distinct()
species_code <- 
  x %>% 
  select(SpecCodeType = speccodetype, SpecCode = speccode, Valid_Aphia = valid_aphia, latin) %>% 
  filter(!is.na(latin)) %>% 
  distinct()
```

Secondly we want to get the NSIBT area, split as well into "Roundfish" subareas
```{r, eval = FALSE}
tmp <- tempfile()
download.file("http://gis.ices.dk/Shapefiles/6951716_output.zip",
              destfile = tmp)
tmpdir <- tempdir()
unzip(tmp, exdir = tmpdir)
ns_area <- rgdal::readOGR(paste0(tmpdir, "/Shapefile.dbf"), verbose = FALSE)
```


Have two options to store retrieved data:

* Export the data as a text file (this way, we could access the data via any software and platform)
* Save the data as a R-binary file

The latter is used in the following step, using the function `save` (read the manual for that function, using ?save):

```{r, eval = FALSE}
save(raw_st, raw_le, raw_ag, species_code, ns_area, file = "data/nsibts_raw.rda")
```

If one then were to start a session at a **later** time the easiest (and least confusing for beginners) is to use the function `load` to get the data into the global environment: 
```{r}
attach("data/nsibts_raw.rda")
```

To get an overview of what is in store lets call the `glimpse`-function:

```{r}
glimpse(raw_st)
glimpse(raw_le)
glimpse(raw_ag)
glimpse(species_code)
```

```{r}
knitr::opts_chunk$set(eval = FALSE)
```

# Tidying
___

The number of variables in each of the dataframes may look overwhelming. Fortunately a lot of the variables are redundant given our objectives. The data frame that contains length also does not fall strictly under the umbrella of being tidy. Here we tidy the data a bit, making later processing a little bit easier.

### The station data


```{r, message = FALSE, warning = FALSE, eval = FALSE}
# Turn column names lower case
colnames(raw_st) <- tolower(colnames(raw_st))
tidy_st <- 
  raw_st %>% 
  # create a unique station id
  unite(id, year, quarter, ship, gear, haulno, remove = FALSE) %>%
  # get proper date
  mutate(timeshot = stringr::str_pad(timeshot, width = 4, side = "left", pad = "0"),
         timeshot = paste0(stringr::str_sub(timeshot, 1, 2),
                           ":",
                           stringr::str_sub(timeshot, 3, 4)),
         datetime = lubridate::ymd_hm(paste(year, month, day, timeshot))) %>% 
  # Get roundfish area
  mutate(area = gisland::geo_inside(shootlong, shootlat, ns_area, "AreaName")) %>% 
  # only valid hauls
  filter(haulval == "V",
         # only in roundfish area
         !is.na(area)) %>% 
  # select only "needed" column
  select(id, survey, year, quarter, ship, gear, haulno, country, hauldur,
         shootlat, shootlong, date = datetime, depth, 
         area, subarea = statrec, daynight, datatype) %>% 
  tbl_df()
```

### The length data

Processing:

* Only distinct records (they are all distinct)
* Filter out data were species code, length-code and length-class are undefined
* Set lengths to millimeter. In the raw records:
    - If **lngtcode** is "1" then the **lngtclass** is in centimeters
    - If **lngtcode** is "." or "0" then the **lngclass** is in millimeters
* Standardise the **haul numbers at length**. In the raw data:
    - If **datatype** in the station table is "C" then **hlnoatlngt** has been standardized to 60 minutes haul
    - If **datatype** in the station table is "R" then **hlnoatlngt** has not been standardized to 60 minutes haul
* Get the latin species name from the `sp`-table
* Select only variable that are needed in further processing
    
```{r, message = FALSE, warning = FALSE, eval = FALSE}
colnames(species_code) <- tolower(colnames(species_code))
colnames(raw_le) <- tolower(colnames(raw_le))
tidy_le <-
  raw_le %>%
  distinct() %>%
  filter(!is.na(speccode), !is.na(lngtclass), !is.na(lngtcode)) %>%
  # EINAR - not sure if this is right, done here to test if any difference
  #         (did though not help with respect to discrepancy here below)
  #filter(specval == 1) %>% 
  unite(id, year, quarter, ship, gear, haulno) %>%
  # only stations that are in the station table (north sea ibts)
  #  may be reduntant
  filter(id %in% tidy_st$id) %>% 
  # length class to cm
  mutate(length = ifelse(lngtcode %in% c("1"), lngtclass, lngtclass / 10),
         hlnoatlngt = hlnoatlngt * subfactor) %>% 
  # get the data type and hauldur
  left_join(tidy_st %>% select(id, datatype, hauldur)) %>% 
  # catch per hour
  mutate(n = ifelse(datatype == "R",
                    hlnoatlngt * 60 / hauldur,
                    hlnoatlngt)) %>% 
  # join with latin name
  left_join(species_code) %>%
  # summarise for each length (basically ignore the sex)
  group_by(id, latin, length) %>% 
  summarise(n = sum(n))
```

### The age data

```{r, eval = FALSE}
colnames(raw_ag) <- tolower(colnames(raw_ag))
tidy_ag <- 
  raw_ag %>% 
  unite(id, year, quarter, ship, gear, haulno) %>% 
  # turn everything to cm
  mutate(length = ifelse(!lngtcode %in% c("1"), lngtclass / 10, lngtclass), 
         indwgt = ifelse(indwgt <= 0, NA, indwgt)) %>% 
  left_join(species_code) %>% 
  select(id, latin, length, sex, maturity, age, wgt = indwgt, n = noatalk) 
```

### Save stuff for later use

```{r, eval = FALSE}
save(tidy_st, tidy_le, tidy_ag, species_code, ns_area, file = "data/nsibts_tidy.rda")
```

