---
title: "TCRENV 2017"
author: "ICES Training Course in the R ENVironment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Time and location

__Time__: 2017-08-21 10:00 - 2017-08-25 15:00

__Location__:
```{r, fig.width = 9, fig.height = 3, message = FALSE, echo=FALSE}
library(leaflet)
content <- paste(sep = "<br/>",
  "<b><a href='http://www.ices.dk'>ICES</a></b>",
  "H. C. Andersens Boulevard 44-46",
  "DK 1553 Copenhagen V, Denmark"
)
leaflet() %>% 
  addTiles() %>% 
  setView(lng = 12.575754, lat = 55.671455, zoom = 15) %>%
   addPopups(12.575754, 55.671455, content,
             options = popupOptions(closeButton = FALSE))
```

### Preamble
___

**Prior to the course** please install:

* Latest version of __R__: see [The Comprehensive R Archive Network](https://cran.r-project.org/)
    * If your platform is Windows, also install [Rtools](https://cran.r-project.org/bin/windows/Rtools)

* Latest version of __RStudio__: see [RStudio Desktop](https://www.rstudio.com/products/rstudio/download/)
    * If you are adventurous install the latest [developmental version](https://www.rstudio.com/products/rstudio/download/preview)

For those that want to get a head we recommend the following read:

  * [R for Data Science](http://r4ds.had.co.nz)
  
We taught this course first time in ICES last year. Last years course [site](http://www.hafro.is/~einarhj/education/tcrenv2016) may give a hint of some of the stuff we cover this year, but expect updates, changes in data source used, etc.

### The workplan
___

(Definitively a) Preliminary schedule:

#### Day 1 - Monday

* Introduction:
    - Rstudio and R projects
    - knitr, markdown, document writing (html, pdf,  docx)
    - reproducible analysis using R
* Getting data into R and out of R:
    - from ones own computer (text files, excel, ...)
    - from the web
    - from API's (ICES Webservices)
    - from databases (Ram's online Postgres database)
* The grammar of data and graphics:
    - Introduction to ggplot2
    - Introduction to dplyr

#### Day 2 - Tuesday

* The grammar of data and graphics - continued:
    - Exploratory data analysis and visualization
* Working with strings and dates
* The base R equivalence

#### Day 3 - Wednesday

* GIS in R
    - using ggplot2 and ggmap
* Case example: Abundance and biomass indices from ICES DATRAS data

#### Day 4 - Thursday

* Applied projects

#### Day 5 - Friday

* The fundamentals of functions and package writing
    - Fundamental of functions and documentation
    - Directory structure and a minimal example
    - Version control (git) and social coding (www.github.com)

### Presentation
___

#### [Getting data into R](pre_data.html)

#### [Getting started with ggplot](pre_ggplot.html)

#### [Getting started with dplyr](pre_dplyr.html)

#### [Applied grammar and then some](pre_grammar2.html)

#### [Working with dates](pre_dates.html)

#### [Reproducible research with R](pre_markdown.html)

#### [Statistic in R](pre_statistics.html)

#### [Plotting spatial data](pre_ggplot2.html)

#### ...

### Bits and pieces
___

#### [Accessing fishbase data](b_fishbase.html)

#### [Generating a package](b_packages.html)

#### [Working with strings](b_strings.html)

#### ...

### Potential projects
___

#### [Good environmental status](prj_ges.html)

#### [Logbook data](prj_log.html)

* **TODO**: Revamp or drop

#### [Viewing stock summary statistics](prj_ss.html)

#### ...

### Ad-hoc stuff
___

#### ...

### Background on some data used
___

#### ...

**The source for this space is found [here](https://github.com/fishvice/tcrenv2017).**
