---
title: "TCRENV 2017"
author: "ICES Training Course in the R ENVironment"
output:
  html_document: default
bibliography: mybib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Time and location
___

__Time__: 2017-08-21 10:00 - 2017-08-25 15:00

__Location__:
```{r, fig.width = 9, fig.height = 3, message = FALSE, echo=FALSE, warning=FALSE}
library(leaflet)
content <- paste(sep = "<br/>",
  "<b><a href='http://www.ices.dk'>ICES</a></b>",
  "H. C. Andersens Boulevard 44-46",
  "DK 1553 Copenhagen V, Denmark"
)
leaflet() %>% 
  addTiles() %>% 
  setView(lng = 12.575754, lat = 55.671455, zoom = 15) %>%
   addPopups(12.575754, 55.671455, content,
             options = popupOptions(closeButton = FALSE))
```

### Preamble
___

**Prior to the course** please install:

* Latest version of __R__: see [The Comprehensive R Archive Network](https://cran.r-project.org/)
    * If your platform is Windows, also install [Rtools](https://cran.r-project.org/bin/windows/Rtools)

* Latest version of __RStudio__: see [RStudio Desktop](https://www.rstudio.com/products/rstudio/download/)
    * If you are adventurous install the latest [developmental version](https://www.rstudio.com/products/rstudio/download/preview)

For those that want to get a head we recommend the following read:

  * [R for Data Science](http://r4ds.had.co.nz) [@wickham2016r]

### The workplan
___

(Definitively a) Preliminary schedule:

#### Day 1 - Monday

* [Introduction](pre_intro.html) - eh
* [Getting data into R](pre_data.html) - bte
* [Getting started with ggplot](pre_ggplot.html) - eh

#### Day 2 - Tuesday

* 09:00 - 10:00 - recapitulation of yesterday
* [Getting started with dplyr](pre_dplyr.html) - bte
* [Working with dates](pre_dates.html) - eh
* [Reproducible research with R](pre_markdown.html) - bte


#### Day 3 - Wednesday

* [Plotting spatial data](pre_ggplot2.html) - eh
* [Relational data](pre_joins.html) - bth
* [Exercises using DATRAS data](pre_case.html) - eh
* Work on your own data
    * Start thinking about creating a presentation of your application to the group. See e.g. [this](http://www.hafro.is/~einarhj/education/tcrenv2016/adhoc/Multispecies_IGFS_DATRAS_maps.html) presentation from a "fellow" of last years course.
    * Ask us if you hit a stumbling block, we get paid for this

#### Day 4 - Thursday

* [Statistics in R](pre_statistics.html) - bte
* Work more on your own data and continue to pester us.
* Participants presentation - if anything available.
* [The fundamentals of functions and package writing]

#### Day 5 - Friday

* [Generating a package](b_packages.html)
* 
* Work and presentation of own data, continued ...
* Introduction to version control (git) and social coding (www.github.com)

### Bits and pieces
___

#### [High resolution gridding](b_highresolutiongridding.html)

#### [Points in polygons](b_points_in_polygons.html)

#### [Getting ICES squares on a map](b_ices_squares.html)

#### [Working with strings](b_strings.html)

#### [dplyr and databases](b_databases.html)

#### [Accessing fishbase data](b_fishbase.html)

#### [Tidying DATRAS](p_tidydatras.html) - could be moved to projects

#### [Loops in R](b_loops.html)

#### [ICCAT data and animation example](presentation.html)

### Potential projects
___

#### [Finish the exercise in Plotting spatial data](pre_ggplot2.html)

#### [ggjoy plot](prj_ggjoy.html)

#### [Good environmental status](prj_ges.html)

#### [Viewing stock summary statistics](prj_ss.html)

### References
___

Web page source is found at [https://github.com/fishvice/tcrenv2017](https://github.com/fishvice/tcrenv2017).
