---
title: "TCRENV 2017"
author: "ICES Training Course in the R ENVironment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Time and location

__Time__: 2017-08-21 10:00 - 2017-08-25 15:00

__Location__:
```{r, fig.width = 9, fig.height = 3, message = FALSE, echo=FALSE}
library(leaflet)
content <- paste(sep = "<br/>",
  "<b><a href='http://www.ices.dk'>ICES</a></b>",
  "H. C. Andersens Boulevard 44-46",
  "DK 1553 Copenhagen V, Denmark"
)
leaflet() %>% 
  addTiles() %>% 
  setView(lng = 12.575754, lat = 55.671455, zoom = 15) %>%
   addPopups(12.575754, 55.671455, content,
             options = popupOptions(closeButton = FALSE))
```

### Preamble
___

**Prior to the course** please install:

* Latest version of __R__: see [The Comprehensive R Archive Network](https://cran.r-project.org/)
    * If your platform is Windows, also install [Rtools](https://cran.r-project.org/bin/windows/Rtools)

* Latest version of __RStudio__: see [RStudio Desktop](https://www.rstudio.com/products/rstudio/download/)
    * If you are adventurous install the latest [developmental version](https://www.rstudio.com/products/rstudio/download/preview)

For those that want to get a head we recommend the following read:

  * [R for Data Science](http://r4ds.had.co.nz)
  
We taught this course first time in ICES last year. Last years course [site](http://www.hafro.is/~einarhj/education/tcrenv2016) may give a hint of some of the stuff we cover this year, but expect updates, changes in data source used, etc.

### The workplan
___

(Definitively a) Preliminary schedule:

#### Day 1 - Monday

* [Introduction](pre_intro.html)
* [Reproducible research with R](pre_markdown.html)
* [Getting data into R](pre_data.html)

#### Day 2 - Tuesday

* [Getting started with ggplot](pre_ggplot.html)
* [Getting started with dplyr](pre_dplyr.html)
* [Working with dates](pre_dates.html)

#### Day 3 - Wednesday

* [Case example](pre_grammar2.html)
* [Statistic in R](pre_statistics.html)
* [Plotting spatial data](pre_ggplot10.html)

#### Day 4 - Thursday

* [Case example](pre_grammar2.html), continued ...
* Applied projects

#### Day 5 - Friday

* Applied projects, continued ...
* The fundamentals of functions and package writing
* Introduction to version control (git) and social coding (www.github.com)

### Bits and pieces
___


#### [Generating a package](b_packages.html)

#### [Working with strings](b_strings.html)

#### [dplyr and databases](b_databases.html)

#### [Accessing fishbase data](b_fishbase.html)

#### ...

### Potential projects
___

#### [Good environmental status](prj_ges.html)

#### [Logbook data](prj_log.html)

* **TODO**: Revamp or drop

#### [Viewing stock summary statistics](prj_ss.html)

#### ...

### Ad-hoc stuff
___

#### ...

### Background on some data used
___

#### ...

**The source for this space is found [here](https://github.com/fishvice/tcrenv2017).**
