---
title: "Getting started with ggplot2"
output: 
  html_document:
    fig_height: 4
    fig_width: 8
    highlight: haddock
    theme: united
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

# Preamble
___

#### Needed libraries for this tutorial:
```{r, message = FALSE, warning = FALSE, results='hide'}
library(ggplot2)
theme_set(theme_grey(base_size = 16))
```

#### Getting example data into R
```{r}
d <- read.csv("http://www.hafro.is/~einarhj/data/minke.csv",
              stringsAsFactors = FALSE)
str(d)
```


# ggplot: Key components
___

ggplot has __three__ key components: 

1.  __data__, this must be a `data.frame`

2. A set of aesthetic mappings (`aes`) between variables in the data and 
   visual properties, and 

3. At least one `layer` which describes how to render each observation.


```{r}
ggplot(data = d) +
  aes(x = age, y = length) + 
  layer(geom = "point", stat = "identity", position = "identity")
```

Generally we do not call `layer` directly but use functions starting with `geom_` that are a shortcut calls to the `layer`-function. Hence the above call is normally written as:
```{r, eval = FALSE}
ggplot(data = d, aes(x = age, y = length)) + geom_point()
```

Different syntax, equivalent outcome:

```{r, eval = FALSE}
ggplot(d, aes(age, length)) + geom_point()
ggplot()                    + geom_point(data = d, aes(age, length))
ggplot(data = d)            + geom_point(aes(x = age, y = length))
ggplot(d)                   + geom_point(aes(age, length))
```

Can be stored as an object for later use:

```{r}
p <- ggplot(d, aes(age, length)) + geom_point()
```

The class:
```{r}
class(p)
```
The structure (a bit of Latin - not run here):
```{r, eval = FALSE}
str(p)
```

# aesthetic
___

Adding more variables to a two dimensional scatterplot can be done by mapping the variables to an aesthetic (colour, fill, size, shape, alpha)

### colour

```{r, out.width = "50%", fig.show = "hold"}
p <- ggplot(d, aes(age, length))
p + geom_point(aes(colour = sex))
p + geom_point(aes(colour = area))
```

Manual control of colours or other palette schemes (here brewer):
```{r, out.width = "50%", fig.show = "hold"}
p + geom_point(aes(colour = sex)) +
  scale_colour_manual(values = c("orange","brown"))
p + geom_point(aes(colour = area)) +
  scale_colour_brewer(palette = "Set1")
```

### shape

```{r}
p + geom_point(aes(age, length, shape = sex))
```

### size

```{r}
p + geom_point(aes(age, length, size = stomach.volume))
```

One can also "fix" the aesthetic manually, e.g.:

```{r}
ggplot(d, aes(age, length)) + geom_point(colour = "blue", shape = 8, size = 10)
```

Note here that the call to colour, shape, etc. is done outside the `aes`-call. One can also combine calls inside and outside the `aes`-function (here we showing overlay of adjacent datapoints):
```{r}
p + geom_point(aes(age, length, size = stomach.volume), alpha = 0.3, col = "red")
```

#### Exercise

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

Create a code that results in these plots:

```{r, echo = FALSE, out.width = "50%", fig.show = "hold"}
p + geom_point(aes(age, length, colour = maturity), size = 2) +
  scale_color_brewer(palette = "Set1")
p + geom_point(aes(age, length, shape = maturity), size = 2)
```

</div>
</div>


# Facetting
___

Splitting a graph into subsets based on a categorical variable. 

```{r}
ggplot(d) + 
  geom_point(aes(age, length, colour = sex)) + 
  facet_wrap(~ area)
```

Here, the variable area is used to spit the plot depending on the location of the sample ("North" and "South").

One can also split the plot using two variables using the function `facet_grid` (here, a very useless plot):

```{r}
ggplot(d) +
  geom_point(aes(age, length)) +
  facet_grid(year ~ sex)
```

# Adding layers
___

The power of ggplot comes into place when one adds layers on top of other layers.

### Add a smoother to a scatterplot

```{r, out.width = "33%", fig.show = "hold"}
p <- ggplot(d, aes(age, length))
p + geom_point() + geom_smooth()
p + geom_point() + geom_smooth(span = 0.1)
p + geom_point() + geom_smooth(method = "lm")
```

In the rightmost plot a linear regression is fitted (check out ?lm).

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">
Create a code that results in this plot:

```{r, echo = FALSE}
ggplot(d, aes(age, length, colour = sex)) + 
  geom_point() + 
  geom_smooth() +
  facet_wrap(~ area) +
  scale_colour_brewer(palette = "Set1")
```

</div>
</div>

# Statistical summary graphs
___

There are some useful *inbuilt* routines within the ggplot2-packages which allows one to create some simple summary plots of the raw data.

### bar plot

One can create bar graph for discrete data using the `geom_bar`

```{r}
ggplot(d, aes(maturity)) + geom_bar()
```

The graph shows the number of observations we have of each maturity stage. Here the argument for stat in the primitive `layer`function is no longer "identity" but "count":

```{r, eval = FALSE}
ggplot(d) +
  aes(maturity) +
  layer(geom = "bar", stat = "count", position = "identity")
```

So the original data is first transformed behind the scene, before being rendered. A simple way to think about this if we were just to generate a simple table of the counts we would do:
```{r}
table(d$maturity, useNA = "ifany")
```

### histograms

For continuous data one uses the `geom_histogram`-function (left default bin-number, right bindwith specified as 50 cm): 

```{r, out.width = "50%", fig.show = "hold"}
p <- ggplot(d, aes(length))
p + geom_histogram()
p + geom_histogram(binwidth = 50)
```

One can add another variable (left) or better use facet (right):

```{r, out.width = "50%", fig.show = "hold"}
p + geom_histogram(aes(fill = sex))
p + geom_histogram() + facet_wrap(~ sex, ncol = 1)
```

### Frequency polygons

Alternatives to histograms for continuous data are frequency polygons:
```{r, out.width = "50%", fig.show = "hold"}
p + geom_freqpoly(lwd = 1)
p + geom_freqpoly(aes(colour = sex), lwd = 1)
```

### Box-plots

Boxplots, which are more condensed summaries of the data than histograms, are called using `geom_boxplot`. Here two versions of the same graph are used, the one on the left is the default, but on the right we have reordered the maturity variable on the x-axis such that the median value of length increases from left to right:
```{r, out.width = "50%", fig.show = "hold"}
ggplot(d, aes(maturity, length)) + geom_boxplot()
p <- ggplot(d, aes(reorder(maturity, length), length)) + geom_boxplot()
p
```

It is sometimes useful to plot the "raw" data over summary plots. Using `geom_point` as an overlay of the above plot (below, left) is not very useful, `geom_jitter` comes here to the rescue (below, right):
```{r, out.width = "50%", fig.show = "hold"}
p + geom_point(colour = "red", alpha = 0.5, size = 1)
p + geom_jitter(colour = "red", alpha = 0.5, size = 1)
```

### Other statistical summaries

Using `stat_summary` one can call specific summary statistics. Here are examples of 4 plots, going from top-left to bottom right we have:

* Raw data with median length at age (red) superimposed
* A pointrange plot showing the mean and the range
* A pointrange plot showing the mean and the standard error
* A pointrange plot showing the bootstrap mean and standard error

```{r, out.width = "50%", fig.show = "hold"}
d$age <- round(d$age)
p <- ggplot(d, aes(age, length))
p + geom_point(alpha = 0.25) + stat_summary(fun.y = "median", geom = "point", colour = "red")
p + stat_summary(fun.y = "mean", fun.ymin = "min", fun.ymax = "max", geom = "pointrange")
p + stat_summary(fun.data = "mean_se")
p + stat_summary(fun.data = "mean_cl_boot")
```

# Some controls
___

### labels

```{r}
p <- ggplot(d, aes(age, length, colour = sex)) + geom_point()
p <- p + labs(x = "Age [year]", y = "Length [cm]", 
         colour = "Sex", 
         title = "My minke plot",
         subtitle = "My nice subtitle")
p
```

### Legend position

```{r, out.width = "50%", fig.show = "hold"}
p + theme(legend.position = "none")
p <- p + theme(legend.position = c(0.8, 0.3))
p
```

### breaks

Controls which values appear as tick marks

```{r, out.width = "50%", fig.show = "hold"}
p + scale_x_continuous(breaks = c(5, 10, 15, 20, 25, 30, 35, 40, 45))
p +
  scale_x_continuous(breaks = seq(5, 45, by = 5)) +
  scale_y_continuous(breaks = seq(500, 950, by = 50))
```

### limits

```{r, out.width = "33%", fig.show = "hold"}
p <- ggplot(d, aes(maturity, length))
p + geom_jitter()
p + geom_jitter() + ylim(600, 800)
p + geom_jitter() + ylim(NA, 800) # setting only upper limit
```

For discrete variables:

```{r}
p + geom_jitter() + ylim(600,800) + xlim("immature","mature")
```

#### warning

But be careful when using lim with summary statistics, e.g.:

```{r, out.width = "50%", fig.show = "hold"}
p + geom_boxplot()
p + geom_boxplot() + ylim(600, 800)
```

Remedy:

```{r}
p + geom_boxplot() + coord_cartesian(ylim = c(600, 800))
```

# Further readings
___

* The ggplot2 site: http://docs.ggplot2.org/current
* The ggplot2 book in the making: https://github.com/hadley/ggplot2-book
    - A rendered version of the book: http://www.hafro.is/~einarhj/education/ggplot2
        - needs to be updates
* R graphics cookbook: http://www.cookbook-r.com/Graphs/

