---
title: "Plotting spatial data"
output: 
  html_document:
    fig_height: 4
    fig_width: 8
    highlight: haddock
    theme: united
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, message = FALSE, warning = FALSE, results='hide'}
library(devtools)
# install_github("tidyverse/ggplot2")
library(geo)
library(maps)
library(mapdata)
library(lubridate)
library(tidyverse)
library(cowplot)
theme_set(theme_grey())
library(ggmap)
minke <- read_csv("data/minke.csv")
```

# Getting started

The minke data has two columns that refer to geographical coordinates. Lets plot the location of the minke sample: 
```{r}
minke %>% 
  ggplot(aes(lon, lat)) +
  geom_point(colour = "red")
```

In the above plot, we basically have mapped the longitude on the x-axis and the latitude on the y-axis. There are two things missing:

* A background, giving the reader a better indication of the geographical region of the sample location.
* The projection (aspect ratio between the x- and the y-axis) looks wrong

## Some backgrounds

* Maps as background for r-plot can come from myriad of sources. Here we take an example of objects available in the map- and mapdata-packages.
* To get the data into ggplot2 friendly form (data.frame) we use the `map_data` function.

```{r}
iceland <- map_data("world", region = "Iceland")
glimpse(iceland)
```

Here we have just a simple dataframe with coordinates and some other variables. We can try map these coordinates to different layers:
```{r}
p <- ggplot(iceland, aes(long, lat)) + labs(x = NULL, y = NULL)
p1 <- p + geom_point()
p2 <- p + geom_line()
p3 <- p + geom_path()
p4 <- p + geom_polygon()
plot_grid(p1, p2, p3, p4)
```

The above sweep of plots demonstrate that background maps are just a set of longitudinal and latitudinal data that are **ordered** in a specific way (check help file for `geom_line` vs `geom_path`

## Projections

As noted above a map is just a xy-plot but with a certain projections. We could try to guess the projections (i.e. the aspect ration of the plot) as done on the left, or better still use the `geom_quickmap` function (on the right):

```{r}
p1 <- p3 + coord_fixed(ratio = 2.4)
p2 <- p3 + coord_quickmap()
plot_grid(p1, p2)
```

The above demonstrates that a spatial map in its simplest term is just an xy-plot with a specific projection. Note that the `coord_quickmap` is only an approximation, if one is operating on a fine scale `coord_map` may be more accurate (actually all maps are wrong when put on a two dimentional pane).

## Selecting a background by boundaries

Instead of selecting a specific named region of a map one can also specify the boundaries, E.g. to get a map of the North Sea:

```{r}
xlim <- c(-5, 12)
ylim <- c(52, 60)
m <- map_data("worldHires", xlim = xlim, ylim = ylim)
ggplot(m, aes(long, lat, group = group)) +
  geom_polygon() +
  coord_quickmap(xlim = xlim, ylim = ylim, expand = FALSE)
```

Here there are two additional element introduced:

* The **variable** group in the m-dataframe is a unique identifier of each separate shape (islands etc.). By specifying the **arguement** group in the `aes`-function one prevents that polygons are drawn across separate elements (try omitting the group-argument).
* The limits are specified inside the call in function `coord_quickmap`. This is because the function `map_data` returns the whole regions that fall within the boundaries (try `coord_quickmap` without any argument).    




#### Exercise

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

1. Play around by plotting different regions or areas
2. Read the `coord_map` help file for projections other than the default "mercator". Try to create a map that looks something along:

```{r, echo = FALSE}
map_data("world") %>% 
  ggplot(aes(x = long, y = lat, group = group)) +
  theme_bw() +
  geom_polygon(fill = "grey") +
  scale_y_continuous(NULL) +
  scale_x_continuous(NULL) +
  coord_map("ortho", xlim = c(-45,30), ylim = c(35,90))
```

</div>
</div>

## Overlay data on maps

Now lets plot our minke data over a map:

1. First we generate a background map:
```{r}
m <- 
  ggplot() +
  theme_bw() +
  geom_polygon(data = iceland, aes(long, lat, group = group), fill = "grey90") +
  coord_map() +
  labs(x = NULL, y = NULL)
```

2. Now add the minke data as a layer:

```{r}
p1 <- m + geom_point(data = minke, aes(lon, lat, colour = sex))
p2 <- m + 
  geom_point(data = minke, aes(lon, lat, size = stomach.weight), colour = "red", alpha = 0.5) +
  scale_size_area(max_size = 10)
plot_grid(p1, p2)
```

# Other background

## Depth contours

In oceanography one is often interested in indicating depth contours. Global relief models from the ETOPO1 dataset hosted on a NOAA server can be accessed using the `getNOAA.bathy`-function in the marmap-package. To access them one specifies the boundary of the data of interest and then, since we are using ggplot for mapping are turned into a data frame using the `fortify`-function:

```{r}
xlim <- c(-28, -10)
ylim <- c(62.5, 67.5)
library(marmap)
depth <- 
  getNOAA.bathy(lon1 = xlim[1], lon2 = xlim[2], lat1 = ylim[1], lat2 = ylim[2], resolution = 1) %>% 
  fortify()
glimpse(depth)
```

So this data is just a set of x (longitude), y (latitudes) and z (depth). The dataset is a raster-grid which we can visualize by using the `geom_raster`-function:

```{r}
depth %>% 
  filter(z <= 0) %>% 
  ggplot() +
  geom_raster(aes(x, y, fill = z))
```

We generate the base map with contours as follows:
```{r}
m <- ggplot() +
  theme_bw() +
  geom_contour(data = depth, aes(x, y, z = z),
               breaks=c(-25, -50, -100, -200, -400),
               colour="black", size=0.1) +
  geom_polygon(data = iceland, aes(long, lat, group = group), fill = "grey") +
  coord_quickmap(xlim = xlim, ylim = ylim, expand = FALSE) +
  labs(x = NULL, y = NULL)
```

Here we have specified depth contour values of 25, 50, 100, 200 and 400 meters. Now we are ready to add the minke data or any other data of interest:
```{r}
m + geom_point(data = minke, aes(lon, lat), colour = "red")
```

## Background images

The ggmap-packages allows one to obtain background images from the web via the `get_map`-function:
```{r}
m2 <- get_map(location = c(-19,65), zoom= 6)
m2 <- ggmap(m2) +
  labs(x = NULL, y = NULL)
m2
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

1. Read the help file for `get_map` and explore some different maptypes and sources available.
2. Overlay the contours just generated on top of such a map
3. Overlay the minke data


```{r, echo = FALSE, eval = FALSE}
m2 + 
  geom_contour(data = depth, aes(x, y, z = z),
               breaks=c(-50, -100, -200, -400),
               colour="white", size=0.2) +
  geom_point(data = minke, aes(lon, lat), colour = "red")
```

</div>
</div>


## Shapefiles

Myriads of useful shapefiles are available on the web, include on the ices web. An example is the shapefile that contains the 10 roundish areas of the North Sea. We can obtain it directly from within R via:

* First we use the `download.file` function, specifying the web location and the destination of where we want to store the file.
* Unzip the file using `unzip` function specifying where we want the content of the zipped file to be stored. 

```{r, eval = FALSE}
download.file(url = "http://gis.ices.dk/Shapefiles/6951716_output.zip",
              destfile = "data/6951716_output.zip", mode = "wb")
unzip("data/6951716_output.zip", exdir = "data")
```

If you now look into the data directory you will have four files starting with "Shapefile" and then four different suffixes. These are standard ESRI files that can be read into R via myriads of functions. Lets use the `st_read` function from the sf-package (which is currently in the process of development):
```{r}
library(sf)
ns_area <- sf::st_read("data/Shapefile.dbf", quiet = TRUE)
class(ns_area)
```

We observe that this is a specialized dataframe. Lets take a peek:
```{r}
ns_area
```

It is beyond the scope of this course to dig deeply into spatial data, but a simple visualization can be obtained via:

```{r}
ns_area %>% 
  mutate(AreaName = as.integer(AreaName)) %>% 
  ggplot() +
  theme_bw() +
  geom_sf(aes(fill = factor(AreaName))) +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "North Sea roundfish area",
       subtitle = "Used for generation of age-length-keys",
       fill = "ALK areas") 
```

# Gridding data

We are often interested in summarizing spatial data onto a grid for presentations. A typical example are logbook data which often can hold millions of records. Lets read in some bottom trawl logbook data from Iceland:
```{r}
# You have to download the data from our sharepoint 
station  <- read_csv("data/logbook_station.csv")
glimpse(station)
```

So we have some 100+ thousand tows with information of towtime (in minutes).

As an example we could summarize the effort by year and ICES statistical rectangle as follows (read the help file on `d2ir`):
```{r}
st <- 
  station %>% 
  mutate(year = year(date),
         sq = geo::d2ir(lat, lon)) %>% 
  group_by(year, sq) %>% 
  summarize(effort = sum(towtime / 60, na.rm = TRUE))
glimpse(st)
```

We have condensed the data to some 292 records but in doing so we have lost the information on the coordinates. Hence before we proceed with any plotting we need to get the central position of the ICES rectangles (using the inverse of the function above):
```{r}
st %>% 
  mutate(lon = geo::ir2d(sq)$lon,     # get the central position of
         lat = geo::ir2d(sq)$lat) %>% #   the ices rectangle
  ggplot() +
  theme_bw() +
  geom_raster(aes(lon, lat, fill = effort/1e3)) +  
  geom_polygon(data = iceland, aes(long, lat, group = group), fill = "grey") +
  coord_quickmap() +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  scale_fill_continuous(type = "viridis", option = "B", direction = -1) +
  labs(fill = "Effort",
       title = "Bottom trawl effort in 1000 hours") +
  facet_wrap(~ year)
```


<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

In addition to the station (tow) file there is an associated file containing information on the catches (kg) by species in each tow (logbook_catch.csv)

1. Read in the file (you download it from the sharepoint)
2. Select two species, e.g. *S. mentella* (61) and plaice (23), summarize the catch of each by year and ICES rectangle and plot it.
3. Add some contours, e.g. 250 and 1000 meters depth to the plot


```{r, echo = FALSE, eval = FALSE}
catch <- read_csv("data/logbook_catch.csv")
station %>% 
  mutate(year = year(date),
         sq = geo::d2ir(lat, lon)) %>% 
  left_join(catch %>% filter(species %in% c(23, 61))) %>% 
  group_by(year, sq, species) %>% 
  summarise(catch = sum(catch, na.rm = TRUE) / 1e6) %>% 
  filter(catch > 0) %>% 
  mutate(lon = geo::ir2d(sq)$lon,     # get the central position of
         lat = geo::ir2d(sq)$lat) %>% #   the ices rectangle
  ggplot() +
  theme_bw() +
  geom_raster(aes(lon, lat, fill = catch)) +  
  geom_contour(data = depth, aes(x, y, z = z),
               breaks=c(-300, -1000),
               colour = "blue", size=0.2) +
  geom_polygon(data = iceland, aes(long, lat, group = group), fill = "grey") +
  coord_quickmap() +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  scale_fill_continuous(type = "viridis", option = "B", direction = -1) +
  labs(fill = "Catch",
       title = "Catch in kt") +
  facet_grid(species ~ year)
```

</div>
</div>

