---
title: "Exercises using DATRAS data"
output: 
  html_document:
    fig_height: 4
    fig_width: 8
    highlight: haddock
    theme: united
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

# Preamble
___

* Here we want to enhance our skills in the use of grammar of data and graphics
* To that end we use the [DATRAS survey data](http://www.ices.dk/marine-data/data-portals/Pages/DATRAS.aspx) as a **case example**.
    * They were retrieved using functions from the icesDatras-package and then tidyed up a bit ([see](datras_importing_and_tidying.html)). The base data is similar to cpue_per_length_per_haul that is one of the [Datras service product](https://datras.ices.dk/Data_products/Download/Download_Data_public.aspx)
* The coding drill below is by no means meant to mimic the "offical" calculation of the DATRAS data, more just to illustrate how one can get an quick overview of the data.

# The data
___

For those that do not already know, procedure in a groundfish surveys goes quite often like this:

* For each haul, all species are identified and the total numbers/weight caught is recorded.
* Length measurement, sex and maturity of a species is generally based on a subsample of the catch.
* Age samples (something we are not going to worry about here) are quite often based on a smaller subsample than sampled for length.

Given the above the data are often stored in separate tables. Here we have three tables.

#### Station table

Dataframe **tidy_st**, it contains the following fields:

  * __id__: Unique station id which for the DATRAS data is a combination of the variables year, quarter, ship, gear and haulno.
  * __survey__: Survey identification (in the case example "NS-IBTS")
  * __quarter__: The quarter in which the survey is conducted
  * __ship__: Ship identification code
  * __gear__: Gear code
  * __haulno__: XXX
  * __hauldur__: Duration of haul in minutes
  * __shootlat__: Latitude of the haul start location
  * __shootlong__: Longitude of the haul start location
  * __datetime__: Haul start time
  * __depth__: Bottom depth in meters
  * __area__: Fish region (in NS-IBTS one has 10 regions)
  * __subarea__: ICES statistical rectangle
  * __daynight__: Categorization code, "D" or "N"
  * __datatype__: Represents is data in the length table has been standardized to 30 minutes in the length table (C or R).
  

#### Length table:

Dataframe **tidy_le**, it contains the following fields:

* __id__: Unique station id (see above)
* __latin__: Species latin name
* __sex__: Sex code
    * __B__: ...
    * __F__: Female
    * __M__: Male
    * __U__: ...
    * __NA__: Not recorded
* __length__: Length in centimeters
* __n__: Number of fish by standardized 30 minutes haul duration
* ... a bunch of additional fields that are redundant


#### Needed packages

```{r}
library(maps)
library(mapdata)
library(tidyverse)
theme_set(theme_grey(base_size = 16))
```

**Recall the dplyr verbs (functions)**:

* __filter__: keep rows matching criteria
* __select__: pick columns by name
* __arrange__: order the rows according to a variable
* __mutate__: add new variables
* __summarise__: reduce variables to values
* __group_by__: Group data into rows
* __left_join__: return all rows from x, and all columns from x and y
* __right_join__: return all rows from y, and all columns from x and y

#### Load the data

Download the data - do this only once:
```{r, eval = FALSE}
download.file(url = "http://www.hafro.is/~einarhj/data/nsibts_tidy.rda",
              destfile = "data/nsibts_tidy.rda")
```

Load the data for the current session:
```{r}
load("data/nsibts_tidy.rda")
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

1. Familiarise yourself and check the structure of the data, such as number of rows, the type of the variables, how many and what years, quarters, species, etc. are in the data set. Check also for missing values.
2. Calculate and plot the number of stations taken per year
3. Pick a year and plot the station locations.

```{r, echo = FALSE, eval = FALSE}
tidy_st %>% 
  group_by(year) %>% 
  summarise(n = n()) %>% 
  ggplot(aes(year, n )) +
  geom_point() +
  geom_line()

tidy_st %>% 
  filter(year == 2017) %>% 
  ggplot(aes(shootlong, shootlat)) +
  geom_point() +
  coord_quickmap() +
  labs(x = NULL, y = NULL)
```

</div>
</div>


# Summarise catch per haul per year
___

Here we are interested in getting information about the catch per unit effort of a particular species over time.

Lets simply generate the median and quantile abundance. We first need to summarise the abundance by each station (basically we ignore/collapse the abundance information by length):

```{r}
le <-
  tidy_le %>% 
  filter(latin == "Gadus morhua") %>% 
  group_by(id) %>% 
  summarise(n = sum(n))
```

The number of unique stations in the length and the station table are:
```{r}
nrow(le)
nrow(tidy_st)
```

So we have missing stations in the length table - this basically means we have a lot a stations where no cod was caught. So we need to join the length data to the station data:

```{r}
d <-
  tidy_st %>% 
  # lets only take some relevant variable
  select(id, year, lon = shootlong, lat = shootlat, sq = subarea) %>% 
  left_join(le)
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

1. Are the number of records in the data.frame **d** as expected, given the number of records in the **tidy_st** data.frame?
2. What does NA in the variable **n** stand for?
3. How many NA's do we have?
4. Can you think of a way replace the NA's with a proper value (hint check the document for the `ifelse`-function)

```{r, echo = FALSE}
# table(is.na(d$n))
d <-
  tidy_st %>% 
  # lets only take some relevant variable
  select(id, year, lon = shootlong, lat = shootlat, sq = subarea) %>% 
  left_join(le) %>% 
  mutate(n = ifelse(is.na(n), 0, n))
```

</div>
</div>


Let's calculate some summary statistics over the years:

```{r}
d.sum <-
  d %>% 
  group_by(year) %>% 
  summarise(n.stations = n(),
            m = mean(n),
            q25 = quantile(n, 0.25),
            q50 = quantile(n, 0.50),
            q75 = quantile(n, 0.75))
```

and plot it:

```{r}
d.sum %>% 
  ggplot(aes(year)) +
  geom_pointrange(aes(y = q50, ymin = q25, ymax = q75))
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

1. Plot the mean and the median on the same graph
2. Standardise both statistics and plot them. Are the abundance trends the same?
3. Modify your code by calculating approximate catch in kg per haul and repeat all the exploration done so far.
4. Try a different species.

```{r, eval = FALSE, echo = FALSE}
d.sum %>% 
  select(year, m, q50) %>% 
  gather(statistics, value, -year) %>% 
  group_by(statistics) %>% 
  mutate(value = value/mean(value)) %>% 
  ggplot(aes(year, value, colour = statistics)) +
  geom_point() +
  geom_line() +
  expand_limits(y = 0)
```

</div>
</div>

# Catch distribution
___

Lets start from scratch and repeat the above calculation:
```{r}
st <- 
  tidy_st %>% 
  as_tibble() %>% 
  select(id, year, lon = shootlong, lat = shootlat, sq = subarea)
d <- 
  tidy_le %>% 
  as_tibble() %>% 
  filter(latin == "Gadus morhua") %>% 
  mutate(b = n * 0.01 * length^3 / 1000) %>% 
  group_by(id) %>% 
  summarise(n = sum(n),
            b = sum(b)) %>% 
  right_join(st) %>% 
  mutate(n = ifelse(is.na(n), 0, n),
         b = ifelse(is.na(b), 0, b))
```

Let's select a year of interest and create a map of the catch per haul:

```{r}
d %>% 
  filter(year %in% c(2005)) %>% 
  ggplot(aes(lon, lat, size = b)) +
  geom_point() +
  coord_quickmap()
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

1. Given what you have learned in [Plotting spatial data](pre_ggplot2.html), modify the above plot code to make it more informative for the reader.
2. Given what you have learned in [Plotting spatial data](pre_ggplot2.html), create a code to calculate and display mean biomass by ICES statistical rectangle.

```{r, eval = FALSE, echo = FALSE}
xlim <- c(-5, 15)
ylim <- c(49, 63)
m <- map_data("worldHires", xlim = xlim, ylim = ylim)
p <- 
  ggplot(m) +
  theme_bw() +
  geom_polygon(aes(long, lat, group = group), fill = "grey") +
  coord_quickmap(xlim = xlim, ylim = ylim, expand = FALSE) +
  labs(x = NULL, y = NULL)
d.2005 <- d %>% filter(year %in% c(2005))
p +
  geom_point(data = d.2005, aes(lon, lat, size = b), alpha = 0.3, colour = "red") +
  scale_size_area(max_size = 10)
```

</div>
</div>

# Length frequency distributions
___

In the above we have ignored the length information in the data. Lets make some simple summarization by length:

```{r}
d <-
  tidy_le %>% 
  filter(latin == "Melanogrammus aeglefinus") %>% 
  as_tibble() %>% 
  # round fish length to the nearest 0.5 cm
  mutate(length = floor(length) + 0.5) %>% 
  # summarise the abundance by length
  group_by(id, length) %>% 
  summarise(n = sum(n)) %>% 
  ungroup()
```

Lets just tally up the total number of fish caught by year. The information on the year is in the station table so we need to join it to the length table before we do the summation:
```{r}
d <-
  d %>% 
  left_join(tidy_st %>% select(id, year)) %>% 
  group_by(year, length) %>% 
  summarise(n = sum(n) / 1e3)
```

Lets look at some selected years:
```{r, fig.height = 6}
d %>% 
  filter(year %in% c(1999:2005)) %>% 
  ggplot(aes(length, n)) +
  geom_col() +
  facet_grid(year ~ .)
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

1. Modify the above code such that it represents distribution of biomass by length.
2. From the distribution by length create summary statistics by year, similar as we did at the beginning of the drill.

```{r, eval = FALSE, echo = FALSE}
tidy_le %>% 
  filter(latin == "Melanogrammus aeglefinus") %>% 
  as_tibble() %>% 
  # round fish length to the nearest 0.5 cm
  mutate(length = floor(length) + 0.5) %>% 
  # summarise the abundance by length
  group_by(id, length) %>% 
  summarise(n = sum(n)) %>% 
  ungroup() %>% 
  mutate(b = n * 0.01 * length^3 / 1000) %>% 
  left_join(tidy_st %>% select(id, year)) %>% 
  group_by(year, length) %>% 
  summarise(b = sum(b) / 1e3) %>% 
  filter(year %in% c(1999:2005)) %>% 
  ggplot(aes(length, b)) +
  geom_col() +
  facet_grid(year ~ .)
```

</div>
</div>