---
title: "ggjoy"
output: 
  html_document
---

This exercise focuses on:

* Installing a new package (ggjoy)
* Reading the help manual for that package or try to get information from the internet on how it operates.
* Trying to create a single pipable code starting from the tidy_ag dataframe and ending with a plot of density distribution of length by age for a selected species that looks something like this:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(ggjoy)
load("data/nsibts_tidy.rda")
```

```{r}
p <-
  tidy_ag %>% 
  filter(latin == "Scomber scombrus", !is.na(age),
         age < 12) %>% 
  select(age, length) %>% 
  drop_na() %>% 
  ggplot(aes(x = length, y = age, group = age, height = ..density..)) +
  geom_joy(alpha = 0.3, fill = "red") +
  scale_y_continuous(breaks = c(1:13)) +
  scale_x_continuous(breaks = seq(10, 45, by = 5)) +
  theme(panel.grid.minor.x = element_line(0),
        panel.grid.minor.y = element_line(0))
p
```

With one singel addition function you can change the plot to this:
```{r}
 p + coord_flip()
```

